---
title: "C. elegans Smaller Gene Set GO Analysis - Manuscript Method"
subtitle: "clusterProfiler over-representation analysis with BH correction"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
    latex_engine: xelatex
geometry: margin=2.5cm
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  out.width = "95%"
)
```

# Purpose

This report reruns the smaller *C. elegans* gene set using the same
over-representation approach used in the manuscript-style analysis:
`clusterProfiler::enrichGO()` with Benjamini-Hochberg (BH) multiple-testing
correction. This is intentionally different from the later `topGO` analysis.
The goal is to document that the smaller gene set does not yield significant
BH-adjusted GO terms under the stricter manuscript method.

# Packages

```{r libraries}
required_pkgs <- c(
  "readxl", "dplyr", "ggplot2", "clusterProfiler", "org.Ce.eg.db",
  "enrichplot", "knitr", "kableExtra"
)

missing_pkgs <- required_pkgs[
  !sapply(required_pkgs, requireNamespace, quietly = TRUE)
]
if (length(missing_pkgs) > 0) {
  BiocManager::install(missing_pkgs, ask = FALSE)
}

invisible(lapply(required_pkgs, library, character.only = TRUE))
```

# Gene Input

```{r load-genes}
file_path_worm <- "./05-2026/list of genes_resubmission.xlsx"
worm_df <- readxl::read_excel(file_path_worm)

wb_col <- grep("WormBase.*Gene|WBGene", names(worm_df),
               ignore.case = TRUE, value = TRUE)[1]
sym_col <- grep("^Gene$|Gene.*[Ss]ymbol|[Ss]ymbol|gene.*name",
                names(worm_df), ignore.case = TRUE, value = TRUE)[1]

cat("Input file:", file_path_worm, "\n")
cat("Total input genes:", nrow(worm_df), "\n")
cat("WormBase ID column:", wb_col, "\n")
cat("Gene symbol column:", sym_col, "\n")
```

# ID Mapping

```{r id-mapping}
map_to_entrez <- function(ids, from_type) {
  ids <- na.omit(unique(ids))
  if (length(ids) == 0) {
    return(data.frame(ENTREZID = character()))
  }

  suppressMessages(
    clusterProfiler::bitr(
      ids,
      fromType = from_type,
      toType = "ENTREZID",
      OrgDb = org.Ce.eg.db
    )
  )
}

wb_ids <- if (!is.na(wb_col)) na.omit(unique(worm_df[[wb_col]])) else character()
sym_ids <- if (!is.na(sym_col)) na.omit(unique(worm_df[[sym_col]])) else character()

map_wb <- map_to_entrez(wb_ids, "WORMBASE")
map_sym <- map_to_entrez(sym_ids, "SYMBOL")

mapping_summary <- data.frame(
  `ID type` = c("WormBase (WORMBASE)", "Gene Symbol (SYMBOL)"),
  `Input genes` = c(length(wb_ids), length(sym_ids)),
  `Mapped Entrez IDs` = c(
    length(unique(map_wb$ENTREZID)),
    length(unique(map_sym$ENTREZID))
  ),
  check.names = FALSE
)

knitr::kable(
  mapping_summary,
  booktabs = TRUE,
  caption = "Entrez ID mapping summary for the smaller gene set"
) |>
  kableExtra::kable_styling(
    latex_options = c("hold_position", "striped"),
    font_size = 10
  )

if (length(unique(map_sym$ENTREZID)) > length(unique(map_wb$ENTREZID))) {
  worm_entrez <- unique(map_sym$ENTREZID)
  id_method <- "Gene Symbol -> Entrez"
} else {
  worm_entrez <- unique(map_wb$ENTREZID)
  id_method <- "WormBase ID -> Entrez"
}

cat("Selected mapping method:", id_method, "\n")
cat("Mapped query genes:", length(worm_entrez), "\n")
```

# Manuscript-Style GO Enrichment

GO over-representation analysis was performed using `enrichGO()` with BH
multiple-testing correction. All terms were retained for reporting and plotting,
but terms were considered significant only if they met the adjusted threshold
used in the manuscript-style workflow.

```{r run-enrichment}
run_go <- function(ontology) {
  clusterProfiler::enrichGO(
    gene = worm_entrez,
    OrgDb = org.Ce.eg.db,
    keyType = "ENTREZID",
    ont = ontology,
    pAdjustMethod = "BH",
    pvalueCutoff = 1,
    qvalueCutoff = 1,
    minGSSize = 3,
    readable = TRUE
  )
}

ego_ce_bp <- run_go("BP")
ego_ce_mf <- run_go("MF")
ego_ce_cc <- run_go("CC")

bp_all <- ego_ce_bp@result
mf_all <- ego_ce_mf@result
cc_all <- ego_ce_cc@result

result_summary <- data.frame(
  Ontology = c("Biological Process", "Molecular Function", "Cellular Component"),
  `Terms tested/returned` = c(nrow(bp_all), nrow(mf_all), nrow(cc_all)),
  `Nominal raw p < 0.05` = c(
    sum(bp_all$pvalue < 0.05, na.rm = TRUE),
    sum(mf_all$pvalue < 0.05, na.rm = TRUE),
    sum(cc_all$pvalue < 0.05, na.rm = TRUE)
  ),
  `Terms with BH adjusted p < 0.05` = c(
    sum(bp_all$p.adjust < 0.05, na.rm = TRUE),
    sum(mf_all$p.adjust < 0.05, na.rm = TRUE),
    sum(cc_all$p.adjust < 0.05, na.rm = TRUE)
  ),
  `Terms with q < 0.2` = c(
    sum(bp_all$qvalue < 0.2, na.rm = TRUE),
    sum(mf_all$qvalue < 0.2, na.rm = TRUE),
    sum(cc_all$qvalue < 0.2, na.rm = TRUE)
  ),
  check.names = FALSE
)

knitr::kable(
  result_summary,
  booktabs = TRUE,
  caption = "Summary of manuscript-style GO enrichment results"
) |>
  kableExtra::kable_styling(
    latex_options = c("hold_position", "striped"),
    font_size = 10
  )
```

## Biological Process

```{r bp-results}
if (nrow(bp_all) == 0) {
  cat("No Biological Process GO terms were significant after BH correction.")
} else {
  bp_show <- bp_all |>
    dplyr::arrange(pvalue) |>
    dplyr::slice_head(n = 20) |>
    dplyr::select(
      ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, qvalue, geneID, Count
    )

  knitr::kable(
    bp_show,
    booktabs = TRUE,
    digits = 4,
    caption = "C. elegans Biological Process GO results using manuscript method"
  ) |>
    kableExtra::kable_styling(
      latex_options = c("scale_down", "hold_position", "striped"),
      font_size = 8
    )
}
```

```{r bp-plot, echo=FALSE}
ego_ce_bp_nominal <- ego_ce_bp
ego_ce_bp_nominal@result <- bp_all |>
  dplyr::filter(pvalue < 0.05) |>
  dplyr::arrange(pvalue)
ego_ce_bp_nominal@pvalueCutoff <- 1
ego_ce_bp_nominal@qvalueCutoff <- 1

if (nrow(ego_ce_bp_nominal@result) > 0) {
  enrichplot::dotplot(
    ego_ce_bp_nominal,
    showCategory = min(20, nrow(ego_ce_bp_nominal@result)),
    font.size = 8,
    color = "pvalue",
    orderBy = "pvalue"
  ) +
    ggplot2::ggtitle("C. elegans GO Enrichment - Biological Process") +
    ggplot2::labs(
      caption = "Exploratory nominal plot: raw p < 0.05 terms are shown; no terms pass BH-adjusted significance."
    )
} else {
  plot.new()
  text(0.5, 0.5, "No nominal BP terms with raw p < 0.05")
}
```

## Molecular Function

```{r mf-results}
if (nrow(mf_all) == 0) {
  cat("No Molecular Function GO terms were significant after BH correction.")
} else {
  mf_show <- mf_all |>
    dplyr::arrange(pvalue) |>
    dplyr::slice_head(n = 20) |>
    dplyr::select(
      ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, qvalue, geneID, Count
    )

  knitr::kable(
    mf_show,
    booktabs = TRUE,
    digits = 4,
    caption = "C. elegans Molecular Function GO results using manuscript method"
  ) |>
    kableExtra::kable_styling(
      latex_options = c("scale_down", "hold_position", "striped"),
      font_size = 8
    )
}
```

```{r mf-plot, echo=FALSE}
ego_ce_mf_nominal <- ego_ce_mf
ego_ce_mf_nominal@result <- mf_all |>
  dplyr::filter(pvalue < 0.05) |>
  dplyr::arrange(pvalue)
ego_ce_mf_nominal@pvalueCutoff <- 1
ego_ce_mf_nominal@qvalueCutoff <- 1

if (nrow(ego_ce_mf_nominal@result) > 0) {
  enrichplot::dotplot(
    ego_ce_mf_nominal,
    showCategory = min(20, nrow(ego_ce_mf_nominal@result)),
    font.size = 8,
    color = "pvalue",
    orderBy = "pvalue"
  ) +
    ggplot2::ggtitle("C. elegans GO Enrichment - Molecular Function") +
    ggplot2::labs(
      caption = "Exploratory nominal plot: raw p < 0.05 terms are shown; no terms pass BH-adjusted significance."
    )
} else {
  plot.new()
  text(0.5, 0.5, "No nominal MF terms with raw p < 0.05")
}
```

## Cellular Component

```{r cc-results}
if (nrow(cc_all) == 0) {
  cat("No Cellular Component GO terms were significant after BH correction.")
} else {
  cc_show <- cc_all |>
    dplyr::arrange(pvalue) |>
    dplyr::slice_head(n = 20) |>
    dplyr::select(
      ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, qvalue, geneID, Count
    )

  knitr::kable(
    cc_show,
    booktabs = TRUE,
    digits = 4,
    caption = "C. elegans Cellular Component GO results using manuscript method"
  ) |>
    kableExtra::kable_styling(
      latex_options = c("scale_down", "hold_position", "striped"),
      font_size = 8
    )
}
```

```{r cc-plot, echo=FALSE}
ego_ce_cc_nominal <- ego_ce_cc
ego_ce_cc_nominal@result <- cc_all |>
  dplyr::filter(pvalue < 0.05) |>
  dplyr::arrange(pvalue)
ego_ce_cc_nominal@pvalueCutoff <- 1
ego_ce_cc_nominal@qvalueCutoff <- 1

if (nrow(ego_ce_cc_nominal@result) > 0) {
  enrichplot::dotplot(
    ego_ce_cc_nominal,
    showCategory = min(20, nrow(ego_ce_cc_nominal@result)),
    font.size = 8,
    color = "pvalue",
    orderBy = "pvalue"
  ) +
    ggplot2::ggtitle("C. elegans GO Enrichment - Cellular Component") +
    ggplot2::labs(
      caption = "Exploratory nominal plot: raw p < 0.05 terms are shown; no terms pass BH-adjusted significance."
    )
} else {
  plot.new()
  text(0.5, 0.5, "No nominal CC terms with raw p < 0.05")
}
```

# Export Results

```{r export-results}
out_dir <- "output_tables_smaller_gene_set_manuscript_method"
dir.create(out_dir, showWarnings = FALSE)

write.csv(bp_all, file.path(out_dir, "Worm_GO_BP_BH_corrected.csv"), row.names = FALSE)
write.csv(mf_all, file.path(out_dir, "Worm_GO_MF_BH_corrected.csv"), row.names = FALSE)
write.csv(cc_all, file.path(out_dir, "Worm_GO_CC_BH_corrected.csv"), row.names = FALSE)
write.csv(result_summary, file.path(out_dir, "GO_BH_corrected_summary.csv"), row.names = FALSE)

cat("Results exported to", out_dir, "\n")
```

# Conclusion

Using the manuscript-style `clusterProfiler::enrichGO()` workflow with BH
correction, the smaller *C. elegans* query set did not produce significant GO
enrichment terms. This supports the interpretation that the lack of significance
is due to the small query size and multiple-testing burden, rather than a failure
of ID mapping. The later `topGO weight01` analysis should therefore be viewed as
a hierarchy-aware follow-up method for extracting directional biological signal
from a small, focused gene list.


