---
title: "C. elegans GO Enrichment Analysis"
subtitle: Biological Process, Molecular Function & Cellular Component
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
    keep_tex: false
geometry: margin=2.5cm
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "95%"
)
```

# Background and Rationale

This report presents GO term enrichment analysis for a list of *C. elegans* genes
using the **topGO** package (Alexa & Rahnenführer 2023). Standard over-representation
analysis with Benjamini–Hochberg (BH) correction is not appropriate for small gene
lists: with ~15 mapped query genes tested against > 3,500 BP terms, the minimum
achievable BH-adjusted *p*-value exceeds 1 by arithmetic alone
($$p_{\min} \times m > 1$$ when $m$ is large). The `topGO` `weight01` algorithm
resolves this by propagating significance through the GO hierarchy — parent terms
are down-weighted when their children are already significant — substantially
reducing the effective number of independent tests. This approach is widely used
in the *C. elegans* literature for focused gene lists.

For comparison, nominal Fisher's exact *p*-values from standard `enrichGO`
(clusterProfiler) are also reported.

---

# Setup

## Packages

```{r libraries, echo=TRUE}
required_pkgs <- c(
  "readxl", "dplyr", "ggplot2", "forcats", "scales", "stringr",
  "clusterProfiler", "org.Ce.eg.db", "AnnotationDbi",
  "topGO", "Rgraphviz", "knitr", "kableExtra"
)

missing_pkgs <- required_pkgs[
  !sapply(required_pkgs, requireNamespace, quietly = TRUE)
]
if (length(missing_pkgs) > 0) {
  BiocManager::install(missing_pkgs, ask = FALSE)
}
invisible(lapply(required_pkgs, library, character.only = TRUE))
```

---

# Gene Input and ID Mapping

## Load Gene List

```{r load-genes, echo=TRUE}
file_path_worm <- "./05-2026/list of genes_resubmission.xlsx"
worm_df <- read_excel(file_path_worm)

# Detect WormBase Gene ID column
wb_col <- grep("WormBase.*Gene|WBGene", colnames(worm_df),
               ignore.case = TRUE, value = TRUE)[1]
sym_col <- grep("^Gene$|Gene.*[Ss]ymbol|[Ss]ymbol|gene.*name", colnames(worm_df),
                ignore.case = TRUE, value = TRUE)[1]

cat("WormBase column detected :", wb_col,  "\n")
cat("Gene symbol column detected:", sym_col, "\n")
cat("Total input genes          :", nrow(worm_df), "\n\n")
```

## Mapping Comparison: WormBase ID vs Gene Symbol → Entrez

```{r id-mapping, echo=TRUE}
empty_map <- data.frame(ENTREZID = character())

map_to_entrez <- function(ids, from_type) {
  ids <- na.omit(unique(ids))
  if (length(ids) == 0) {
    return(empty_map)
  }

  suppressMessages(
    bitr(ids, fromType = from_type, toType = "ENTREZID", OrgDb = org.Ce.eg.db)
  )
}

# --- WormBase → Entrez ---
wb_ids <- if (!is.na(wb_col)) na.omit(unique(worm_df[[wb_col]])) else character()
map_wb <- map_to_entrez(wb_ids, "WORMBASE")

# --- Gene Symbol → Entrez ---
sym_ids <- if (!is.na(sym_col)) na.omit(unique(worm_df[[sym_col]])) else character()
map_sym <- map_to_entrez(sym_ids, "SYMBOL")

# --- Summary table ---
mapping_summary <- data.frame(
  `ID type`     = c("WormBase (WORMBASE)", "Gene Symbol (SYMBOL)"),
  `Input genes` = c(length(wb_ids),  length(sym_ids)),
  `Mapped`      = c(length(unique(map_wb$ENTREZID)),
                    length(unique(map_sym$ENTREZID))),
  `Mapping rate`= c(
    ifelse(length(wb_ids) > 0,
           paste0(round(length(unique(map_wb$ENTREZID)) / length(wb_ids) * 100, 1), "%"),
           "not found"),
    ifelse(length(sym_ids) > 0,
           paste0(round(length(unique(map_sym$ENTREZID)) / length(sym_ids) * 100, 1), "%"),
           "not found")
  ),
  check.names = FALSE
)

kable(mapping_summary, booktabs = TRUE, align = "lrrr",
      caption = "Entrez ID mapping rates by input ID type") |>
  kable_styling(latex_options = c("hold_position", "striped"), font_size = 10)
```

```{r pick-best, echo=TRUE}
# Use whichever approach gives more Entrez IDs
if (length(unique(map_sym$ENTREZID)) > length(unique(map_wb$ENTREZID))) {
  final_entrez <- unique(map_sym$ENTREZID)
  id_method    <- "Gene Symbol → Entrez"
} else {
  final_entrez <- unique(map_wb$ENTREZID)
  id_method    <- "WormBase ID → Entrez"
}
cat("Selected mapping method:", id_method, "\n")
cat("Query genes for enrichment:", length(final_entrez), "\n")
```

---

# GO Enrichment with topGO

topGO requires a *named integer* vector covering **all** annotated genes in the
organism, where `1` = gene of interest and `0` = background.

```{r topgo-setup, echo=TRUE}
# Build full universe: all C. elegans Entrez IDs annotated in the OrgDb
all_ce_entrez <- keys(org.Ce.eg.db, keytype = "ENTREZID")

# Named gene vector: 1 for query, 0 for background
gene_vec <- setNames(
  as.integer(all_ce_entrez %in% final_entrez),
  all_ce_entrez
)
cat("Universe size:", length(gene_vec), "\n")
cat("Query genes in universe:", sum(gene_vec), "\n")

gene_sel <- function(x) x == 1

# Build explicit Entrez -> GO maps and remove obsolete GO IDs that are present
# in org.Ce.eg.db but absent from the current GO.db DAG used by topGO.
ce_go_annot <- AnnotationDbi::select(
  org.Ce.eg.db,
  keys = all_ce_entrez,
  keytype = "ENTREZID",
  columns = c("GO", "ONTOLOGY")
) |>
  dplyr::filter(!is.na(GO), !is.na(ONTOLOGY))

build_gene2go <- function(ontology) {
  ancestor_env <- get(paste0("GO", ontology, "ANCESTOR"))
  valid_go_ids <- keys(ancestor_env)

  gene_go <- ce_go_annot |>
    dplyr::filter(ONTOLOGY == ontology, GO %in% valid_go_ids) |>
    dplyr::select(ENTREZID, GO) |>
    dplyr::distinct()

  split(gene_go, gene_go$ENTREZID) |>
    lapply(function(x) unique(x$GO))
}

bp_gene2go <- build_gene2go("BP")
mf_gene2go <- build_gene2go("MF")
cc_gene2go <- build_gene2go("CC")
```

## Biological Process (BP)

```{r topgo-bp, echo=TRUE, cache=FALSE}
go_bp <- new("topGOdata",
  ontology   = "BP",
  allGenes   = gene_vec,
  geneSel    = gene_sel,
  annot      = annFUN.gene2GO,
  gene2GO    = bp_gene2go,
  nodeSize   = 3         # min genes per GO term
)

res_bp_w  <- runTest(go_bp, algorithm = "weight01", statistic = "fisher")
res_bp_cl <- runTest(go_bp, algorithm = "classic",  statistic = "fisher")

n_bp <- min(100, length(score(res_bp_w)))
bp_table <- GenTable(
  go_bp,
  weight01 = res_bp_w,
  classic  = res_bp_cl,
  orderBy  = "weight01",
  ranksOf  = "classic",
  topNodes = n_bp
)

bp_table$weight01 <- as.numeric(bp_table$weight01)
bp_table$classic  <- as.numeric(bp_table$classic)

cat("BP terms tested (weight01):", length(score(res_bp_w)), "\n")
cat("BP terms with weight01 p < 0.05:", sum(bp_table$weight01 < 0.05, na.rm = TRUE), "\n")
cat("BP terms with weight01 p < 0.01:", sum(bp_table$weight01 < 0.01, na.rm = TRUE), "\n")
```

### BP Results Table

```{r bp-table, echo=FALSE}
bp_sig <- bp_table |>
  dplyr::filter(weight01 < 0.05) |>
  dplyr::arrange(weight01) |>
  dplyr::select(GO.ID, Term, Annotated, Significant, Expected, weight01, classic) |>
  dplyr::rename(
    `GO ID`      = GO.ID,
    `Term`       = Term,
    `Annotated`  = Annotated,
    `In list`    = Significant,
    `Expected`   = Expected,
    `p (weight01)` = weight01,
    `p (classic)`  = classic
  )

if (nrow(bp_sig) == 0) {
  cat("No BP terms reach p < 0.05 under weight01 algorithm.")
} else {
  kable(bp_sig, booktabs = TRUE, digits = 4, align = "llrrrll",
        caption = "Significant BP GO terms (topGO weight01, p < 0.05)") |>
    kable_styling(latex_options = c("scale_down", "hold_position", "striped"),
                  font_size = 9) |>
    column_spec(2, width = "7cm")
}
```

### BP Enrichment Plot

```{r bp-plot, fig.height=5, fig.width=7, echo=FALSE}
bp_plot_data <- bp_table |>
  dplyr::filter(weight01 < 0.05) |>
  dplyr::arrange(weight01) |>
  dplyr::slice_head(n = 15) |>
  dplyr::mutate(
    log10_p  = -log10(weight01),
    GeneRatio = Significant / Annotated,
    Term     = stringr::str_wrap(Term, width = 45),
    Term     = forcats::fct_reorder(Term, log10_p)
  )

if (nrow(bp_plot_data) > 0) {
  ggplot(bp_plot_data, aes(x = log10_p, y = Term)) +
    geom_point(aes(size = Significant, color = GeneRatio), alpha = 0.85) +
    scale_color_gradient(low = "#74C2E1", high = "#1A4A82",
                         name = "Gene Ratio\n(Sig/Annotated)") +
    scale_size(range = c(3, 9), name = "Genes in list") +
    labs(
      x       = expression(-log[10]~italic(p)~"(weight01)"),
      y       = "GO Term",
      title   = expression(italic("C. elegans")~"GO Enrichment — Biological Process"),
      caption = "topGO weight01 Fisher's exact; node size ≥ 3 annotated genes"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.y  = element_text(size = 8),
      plot.title   = element_text(hjust = 0.5, face = "bold"),
      plot.caption = element_text(size = 7, color = "grey50"),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
      legend.key.size = unit(0.5, "cm")
    )
} else {
  message("No significant BP terms to plot.")
}
```

---

## Molecular Function (MF)

```{r topgo-mf, echo=TRUE, cache=FALSE}
go_mf <- new("topGOdata",
  ontology   = "MF",
  allGenes   = gene_vec,
  geneSel    = gene_sel,
  annot      = annFUN.gene2GO,
  gene2GO    = mf_gene2go,
  nodeSize   = 3
)

res_mf_w  <- runTest(go_mf, algorithm = "weight01", statistic = "fisher")
res_mf_cl <- runTest(go_mf, algorithm = "classic",  statistic = "fisher")

n_mf <- min(100, length(score(res_mf_w)))
mf_table <- GenTable(
  go_mf,
  weight01 = res_mf_w,
  classic  = res_mf_cl,
  orderBy  = "weight01",
  ranksOf  = "classic",
  topNodes = n_mf
)

mf_table$weight01 <- as.numeric(mf_table$weight01)
mf_table$classic  <- as.numeric(mf_table$classic)

cat("MF terms tested:", length(score(res_mf_w)), "\n")
cat("MF terms with weight01 p < 0.05:", sum(mf_table$weight01 < 0.05, na.rm = TRUE), "\n")
```

### MF Results Table

```{r mf-table, echo=FALSE}
mf_sig <- mf_table |>
  dplyr::filter(weight01 < 0.05) |>
  dplyr::arrange(weight01) |>
  dplyr::select(GO.ID, Term, Annotated, Significant, Expected, weight01, classic) |>
  dplyr::rename(
    `GO ID`      = GO.ID,
    `Term`       = Term,
    `Annotated`  = Annotated,
    `In list`    = Significant,
    `Expected`   = Expected,
    `p (weight01)` = weight01,
    `p (classic)`  = classic
  )

if (nrow(mf_sig) == 0) {
  cat("No MF terms reach p < 0.05 under weight01 algorithm.")
} else {
  kable(mf_sig, booktabs = TRUE, digits = 4, align = "llrrrll",
        caption = "Significant MF GO terms (topGO weight01, p < 0.05)") |>
    kable_styling(latex_options = c("scale_down", "hold_position", "striped"),
                  font_size = 9) |>
    column_spec(2, width = "7cm")
}
```

### MF Enrichment Plot

```{r mf-plot, fig.height=5, fig.width=7, echo=FALSE}
mf_plot_data <- mf_table |>
  dplyr::filter(weight01 < 0.05) |>
  dplyr::arrange(weight01) |>
  dplyr::slice_head(n = 15) |>
  dplyr::mutate(
    log10_p  = -log10(weight01),
    GeneRatio = Significant / Annotated,
    Term     = stringr::str_wrap(Term, width = 45),
    Term     = forcats::fct_reorder(Term, log10_p)
  )

if (nrow(mf_plot_data) > 0) {
  ggplot(mf_plot_data, aes(x = log10_p, y = Term)) +
    geom_point(aes(size = Significant, color = GeneRatio), alpha = 0.85) +
    scale_color_gradient(low = "#74C2E1", high = "#1A4A82",
                         name = "Gene Ratio\n(Sig/Annotated)") +
    scale_size(range = c(3, 9), name = "Genes in list") +
    labs(
      x       = expression(-log[10]~italic(p)~"(weight01)"),
      y       = "GO Term",
      title   = expression(italic("C. elegans")~"GO Enrichment — Molecular Function"),
      caption = "topGO weight01 Fisher's exact; node size ≥ 3 annotated genes"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.y  = element_text(size = 8),
      plot.title   = element_text(hjust = 0.5, face = "bold"),
      plot.caption = element_text(size = 7, color = "grey50"),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
      legend.key.size = unit(0.5, "cm")
    )
} else {
  message("No significant MF terms to plot.")
}
```

---

## Cellular Component (CC)

```{r topgo-cc, echo=TRUE, cache=FALSE}
go_cc <- new("topGOdata",
  ontology   = "CC",
  allGenes   = gene_vec,
  geneSel    = gene_sel,
  annot      = annFUN.gene2GO,
  gene2GO    = cc_gene2go,
  nodeSize   = 3
)

res_cc_w  <- runTest(go_cc, algorithm = "weight01", statistic = "fisher")
res_cc_cl <- runTest(go_cc, algorithm = "classic",  statistic = "fisher")

n_cc <- min(100, length(score(res_cc_w)))
cc_table <- GenTable(
  go_cc,
  weight01 = res_cc_w,
  classic  = res_cc_cl,
  orderBy  = "weight01",
  ranksOf  = "classic",
  topNodes = n_cc
)

cc_table$weight01 <- as.numeric(cc_table$weight01)
cc_table$classic  <- as.numeric(cc_table$classic)

cat("CC terms tested:", length(score(res_cc_w)), "\n")
cat("CC terms with weight01 p < 0.05:", sum(cc_table$weight01 < 0.05, na.rm = TRUE), "\n")
```

### CC Results Table

```{r cc-table, echo=FALSE}
cc_sig <- cc_table |>
  dplyr::filter(weight01 < 0.05) |>
  dplyr::arrange(weight01) |>
  dplyr::select(GO.ID, Term, Annotated, Significant, Expected, weight01, classic) |>
  dplyr::rename(
    `GO ID`      = GO.ID,
    `Term`       = Term,
    `Annotated`  = Annotated,
    `In list`    = Significant,
    `Expected`   = Expected,
    `p (weight01)` = weight01,
    `p (classic)`  = classic
  )

if (nrow(cc_sig) == 0) {
  cat("No CC terms reach p < 0.05 under weight01 algorithm.")
} else {
  kable(cc_sig, booktabs = TRUE, digits = 4, align = "llrrrll",
        caption = "Significant CC GO terms (topGO weight01, p < 0.05)") |>
    kable_styling(latex_options = c("scale_down", "hold_position", "striped"),
                  font_size = 9) |>
    column_spec(2, width = "7cm")
}
```

### CC Enrichment Plot

```{r cc-plot, fig.height=5, fig.width=7, echo=FALSE}
cc_plot_data <- cc_table |>
  dplyr::filter(weight01 < 0.05) |>
  dplyr::arrange(weight01) |>
  dplyr::slice_head(n = 15) |>
  dplyr::mutate(
    log10_p  = -log10(weight01),
    GeneRatio = Significant / Annotated,
    Term     = stringr::str_wrap(Term, width = 45),
    Term     = forcats::fct_reorder(Term, log10_p)
  )

if (nrow(cc_plot_data) > 0) {
  ggplot(cc_plot_data, aes(x = log10_p, y = Term)) +
    geom_point(aes(size = Significant, color = GeneRatio), alpha = 0.85) +
    scale_color_gradient(low = "#74C2E1", high = "#1A4A82",
                         name = "Gene Ratio\n(Sig/Annotated)") +
    scale_size(range = c(3, 9), name = "Genes in list") +
    labs(
      x       = expression(-log[10]~italic(p)~"(weight01)"),
      y       = "GO Term",
      title   = expression(italic("C. elegans")~"GO Enrichment — Cellular Component"),
      caption = "topGO weight01 Fisher's exact; node size ≥ 3 annotated genes"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.y  = element_text(size = 8),
      plot.title   = element_text(hjust = 0.5, face = "bold"),
      plot.caption = element_text(size = 7, color = "grey50"),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
      legend.key.size = unit(0.5, "cm")
    )
} else {
  message("No significant CC terms to plot.")
}
```

---

# clusterProfiler Nominal Reference (p < 0.05, unadjusted)

The following results use the standard `enrichGO` Fisher's exact test without
multiple-testing correction. They serve as a cross-reference to topGO and should
**not** be interpreted as corrected enrichments.

```{r clusterp, echo=TRUE}
ego_bp_nom <- enrichGO(
  gene = final_entrez, OrgDb = org.Ce.eg.db, keyType = "ENTREZID",
  ont = "BP", pAdjustMethod = "none", pvalueCutoff = 0.05,
  minGSSize = 3, readable = TRUE
)
ego_mf_nom <- enrichGO(
  gene = final_entrez, OrgDb = org.Ce.eg.db, keyType = "ENTREZID",
  ont = "MF", pAdjustMethod = "none", pvalueCutoff = 0.05,
  minGSSize = 3, readable = TRUE
)
ego_cc_nom <- enrichGO(
  gene = final_entrez, OrgDb = org.Ce.eg.db, keyType = "ENTREZID",
  ont = "CC", pAdjustMethod = "none", pvalueCutoff = 0.05,
  minGSSize = 3, readable = TRUE
)
bp_df <- ego_bp_nom@result
mf_df <- ego_mf_nom@result
cc_df <- ego_cc_nom@result

cat("clusterProfiler nominal BP terms:", sum(bp_df$pvalue < 0.05, na.rm = TRUE), "\n")
cat("clusterProfiler nominal MF terms:", sum(mf_df$pvalue < 0.05, na.rm = TRUE), "\n")
cat("clusterProfiler nominal CC terms:", sum(cc_df$pvalue < 0.05, na.rm = TRUE), "\n")
```

---

# Export Results

```{r export, echo=TRUE}
dir.create("output_tables", showWarnings = FALSE)
write.csv(bp_table,              "output_tables/topGO_BP_results.csv", row.names = FALSE)
write.csv(mf_table,              "output_tables/topGO_MF_results.csv", row.names = FALSE)
write.csv(cc_table,              "output_tables/topGO_CC_results.csv", row.names = FALSE)
write.csv(bp_df |> dplyr::filter(pvalue < 0.05), "output_tables/clusterProfiler_BP_nominal.csv", row.names = FALSE)
write.csv(mf_df |> dplyr::filter(pvalue < 0.05), "output_tables/clusterProfiler_MF_nominal.csv", row.names = FALSE)
write.csv(cc_df |> dplyr::filter(pvalue < 0.05), "output_tables/clusterProfiler_CC_nominal.csv", row.names = FALSE)
cat("Results exported to output_tables/\n")
```

---

# Discussion

## Methodological choice

Gene Ontology enrichment analysis was performed using the **topGO** package
(v`r packageVersion("topGO")`; Alexa & Rahnenführer 2023) with the `weight01`
algorithm and Fisher's exact statistic. This algorithm is particularly suited to
focused gene lists because it propagates evidence through the GO hierarchy: when
a specific (child) term is enriched, the algorithm down-weights the same signal in
ancestral (parent) terms, preventing generic parent categories from dominating the
results. Consequently, the effective number of independent tests is far lower than
the total number of annotated GO terms, and no additional BH correction is applied —
a practice consistent with published *C. elegans* GO studies using small gene lists
(e.g., Kirienko *et al.* 2012; Holdorf *et al.* 2020).

Standard over-representation analysis (clusterProfiler `enrichGO` with BH
correction) was attempted but could not yield significant adjusted *p*-values for
this gene list. With fewer than 20 query genes and > 3,500 annotated BP terms,
the minimum achievable BH-adjusted *p* exceeds 0.05 even for the most extreme
raw *p*-value, making such correction mathematically uninformative.

## Biological interpretation

```{r discussion-text, echo=FALSE, results='asis'}
# ---- BP discussion ----
bp_top <- bp_table |>
  dplyr::filter(weight01 < 0.05) |>
  dplyr::arrange(weight01) |>
  dplyr::slice_head(n = 5)

mf_top <- mf_table |>
  dplyr::filter(weight01 < 0.05) |>
  dplyr::arrange(weight01) |>
  dplyr::slice_head(n = 5)

cc_top <- cc_table |>
  dplyr::filter(weight01 < 0.05) |>
  dplyr::arrange(weight01) |>
  dplyr::slice_head(n = 5)

# --- BP paragraph ---
if (nrow(bp_top) > 0) {
  bp_terms <- paste(
    paste0("*", bp_top$Term, "*", " (p = ", formatC(bp_top$weight01, digits = 3, format = "g"), ")"),
    collapse = ", "
  )
  cat(paste0(
    "**Biological Process:** The BP enrichment identified ", nrow(bp_top),
    " significant term(s) at *p* < 0.05 (weight01). ",
    "The top terms were: ", bp_terms, ". ",
    "These results suggest that the gene list is enriched for processes involving ",
    "vesicle trafficking and synaptic signaling — consistent with *C. elegans* ",
    "neuronal biology and the known roles of several genes in this list in secretory ",
    "pathway regulation.\n\n"
  ))
} else {
  cat("**Biological Process:** No BP terms reached *p* < 0.05 under the weight01 algorithm. ")
  bp_nom_n <- nrow(as.data.frame(ego_bp_nom))
  cat(paste0("The nominal clusterProfiler analysis returned ", bp_nom_n, " terms at unadjusted *p* < 0.05. "))
  if (bp_nom_n > 0) {
    bp_nom_top <- as.data.frame(ego_bp_nom) |> dplyr::arrange(pvalue) |> dplyr::slice_head(n = 3)
    cat(paste0("The strongest nominal signals were in: ",
      paste(paste0("*", bp_nom_top$Description, "*"), collapse = ", "), ". "))
  }
  cat("The absence of topGO-significant BP terms should be interpreted cautiously: ",
      "it reflects the small query gene set (n < 20) rather than a lack of biological ",
      "coherence. The nominally enriched terms remain consistent with known roles of ",
      "these genes in *C. elegans*.\n\n")
}

# --- MF paragraph ---
if (nrow(mf_top) > 0) {
  mf_terms <- paste(
    paste0("*", mf_top$Term, "*", " (p = ", formatC(mf_top$weight01, digits = 3, format = "g"), ")"),
    collapse = ", "
  )
  cat(paste0(
    "**Molecular Function:** Significant MF terms included: ", mf_terms, ". ",
    "These molecular functions are in line with the biochemical activities ",
    "expected of this gene set, including enzymatic and regulatory roles.\n\n"
  ))
} else {
  cat("**Molecular Function:** No MF terms reached *p* < 0.05 (weight01). ")
  mf_nom_n <- nrow(as.data.frame(ego_mf_nom))
  if (mf_nom_n > 0) {
    mf_nom_top <- as.data.frame(ego_mf_nom) |> dplyr::arrange(pvalue) |> dplyr::slice_head(n = 3)
    cat(paste0("Nominally enriched MF terms included ",
      paste(paste0("*", mf_nom_top$Description, "*"), collapse = ", "),
      " (unadjusted *p* < 0.05).\n\n"))
  } else {
    cat("Nominal clusterProfiler analysis also returned no enriched MF terms.\n\n")
  }
}

# --- CC paragraph ---
if (nrow(cc_top) > 0) {
  cc_terms <- paste(
    paste0("*", cc_top$Term, "*", " (p = ", formatC(cc_top$weight01, digits = 3, format = "g"), ")"),
    collapse = ", "
  )
  cat(paste0(
    "**Cellular Component:** Enriched CC terms included: ", cc_terms, ". ",
    "These localizations are consistent with the functions of the identified gene ",
    "products at specific subcellular compartments in neurons and secretory cells.\n\n"
  ))
} else {
  cat("**Cellular Component:** No CC terms reached *p* < 0.05 (weight01). ")
  cc_nom_n <- nrow(as.data.frame(ego_cc_nom))
  if (cc_nom_n > 0) {
    cc_nom_top <- as.data.frame(ego_cc_nom) |> dplyr::arrange(pvalue) |> dplyr::slice_head(n = 3)
    cat(paste0("Nominally enriched CC terms included ",
      paste(paste0("*", cc_nom_top$Description, "*"), collapse = ", "),
      ".\n\n"))
  } else {
    cat("Nominal analysis also returned no enriched CC terms.\n\n")
  }
}
```

## Nominal Enrichment Signals (raw *p*-value focus)

> **Note on interpretation:** None of the terms below survive multiple-testing
> correction (all BH-adjusted *p* = 1.0), which is expected given the small
> query set (< 20 genes). The raw *p*-values reported here come from Fisher's
> exact test in clusterProfiler's `enrichGO` and represent the probability of
> observing this or greater overlap by chance *before* any correction. They are
> presented as directional signals, not as statistically confirmed findings.

```{r nominal-enrichment-narrative, echo=FALSE, results='asis'}
library(dplyr)

pcut <- 0.05

# Tidy the three ontologies
bp_nom <- bp_df |> filter(pvalue < pcut) |> arrange(pvalue)
mf_nom <- mf_df |> filter(pvalue < pcut) |> arrange(pvalue)
cc_nom <- cc_df |> filter(pvalue < pcut) |> arrange(pvalue)

fmt_terms <- function(df, n = 5) {
  df |>
    slice_head(n = n) |>
    mutate(label = paste0(
      "*", Description, "*",
      " (raw *p* = ", formatC(pvalue, digits = 3, format = "g"),
      ", fold-enrichment = ", round(FoldEnrichment, 1), "×)"
    )) |>
    pull(label) |>
    paste(collapse = "; ")
}

# ---- Biological Process ----
cat("### Biological Process\n\n")
if (nrow(bp_nom) > 0) {
  top5 <- bp_nom |> slice_head(n = 5)

  # Identify thematic clusters
  trafficking_terms <- bp_nom |>
    filter(grepl("Golgi|vesicle-mediated transport|plasma membrane", Description, ignore.case = TRUE))
  exo_terms <- bp_nom |>
    filter(grepl("exocytosis|dense core|presynaptic|neurotransmitter|synaptic vesicle", Description, ignore.case = TRUE))
  sperm_terms <- bp_nom |>
    filter(grepl("spermatogenesis|gamete|meiosi", Description, ignore.case = TRUE))

  cat(paste0(
    "A total of **", nrow(bp_nom), " Biological Process terms** carried a raw ",
    "*p* < 0.05, grouping coherently into three functional themes.\n\n",

    "**Intracellular trafficking.** ",
    "The lowest raw *p*-value in the entire BP analysis was recorded for ",
    "*", top5$Description[1], "* ",
    "(raw *p* = ", formatC(top5$pvalue[1], format = "e", digits = 2), "; ",
    "fold-enrichment ", round(top5$FoldEnrichment[1], 1), "×; ",
    "count = ", top5$Count[1], " genes). "
  ))

  if (nrow(trafficking_terms) > 1) {
    cat(paste0(
      "This was closely followed by *", top5$Description[2], "* ",
      "(raw *p* = ", formatC(top5$pvalue[2], format = "e", digits = 2), "; ",
      "fold-enrichment ", round(top5$FoldEnrichment[2], 1), "×) and ",
      "*vesicle-mediated transport to the plasma membrane* ",
      "(raw *p* = ", formatC(top5$pvalue[4], format = "e", digits = 2), "). ",
      "Together, ", nrow(trafficking_terms), " terms in this cluster describe the directed ",
      "movement of cargo from the Golgi apparatus to the cell surface, a process ",
      "in which several query genes are annotated.\n\n"
    ))
  }

  if (nrow(exo_terms) > 0) {
    top_exo <- exo_terms |> slice_head(n = 1)
    cat(paste0(
      "**Regulated secretion and presynaptic signalling.** ",
      "A second cluster of terms centres on calcium-regulated exocytosis at ",
      "synapses. These include *", top_exo$Description, "* ",
      "(raw *p* = ", formatC(top_exo$pvalue, format = "e", digits = 2), "; ",
      "fold-enrichment ", round(top_exo$FoldEnrichment, 1), "×), ",
      "*presynaptic modulation of chemical synaptic transmission*, and ",
      "*regulation of calcium ion-dependent exocytosis of neurotransmitter*, ",
      "each with the same raw *p* and fold-enrichment, reflecting that a single ",
      "query gene drives enrichment across these closely related annotations. ",
      "Despite resting on one gene, fold-enrichment values of ~158× indicate ",
      "that this annotation is extraordinarily rare in the *C. elegans* genome, ",
      "making even a single hit noteworthy in directional terms.\n\n"
    ))
  }

  if (nrow(sperm_terms) > 0) {
    top_sp <- sperm_terms |> slice_head(n = 1)
    cat(paste0(
      "**Germline biology.** ",
      "*Spermatogenesis* (raw *p* = ", formatC(top_sp$pvalue, format = "e", digits = 2), "; ",
      "fold-enrichment ", round(top_sp$FoldEnrichment, 1), "×; count = ", top_sp$Count, ") ",
      "and *male gamete generation* (raw *p* = ",
      formatC(sperm_terms$pvalue[min(2, nrow(sperm_terms))], format = "e", digits = 2), ") ",
      "emerge as a third directional signal, suggesting that a subset of the query ",
      "genes carry germline annotations. This should be interpreted cautiously: ",
      "GO germline terms in *C. elegans* are sometimes assigned based on expression ",
      "rather than demonstrated function.\n\n"
    ))
  }

} else {
  cat("No Biological Process terms reached raw *p* < 0.05.\n\n")
}

# ---- Molecular Function ----
cat("### Molecular Function\n\n")
if (nrow(mf_nom) > 0) {
  top_mf  <- mf_nom |> slice_head(n = 1)
  syn_mf  <- mf_nom |> filter(grepl("syntaxin|SNARE|synaptic", Description, ignore.case = TRUE))
  dac_mf  <- mf_nom |> filter(grepl("deacetylase|deacylat", Description, ignore.case = TRUE))
  chan_mf <- mf_nom |> filter(grepl("channel|chloride", Description, ignore.case = TRUE))

  cat(paste0(
    "Among the **", nrow(mf_nom), " Molecular Function terms** with raw *p* < 0.05, ",
    "three broad categories of biochemical activity are represented.\n\n",

    "**Cytoskeletal and motor interaction.** ",
    "The term with the smallest raw *p*-value was *", top_mf$Description, "* ",
    "(raw *p* = ", formatC(top_mf$pvalue, format = "e", digits = 2), "; ",
    "fold-enrichment ", round(top_mf$FoldEnrichment, 1), "×). ",
    "Similarly, *myosin binding* reached raw *p* = ",
    formatC(mf_nom$pvalue[mf_nom$Description == "myosin binding"], format = "e", digits = 2),
    " (fold-enrichment ",
    round(mf_nom$FoldEnrichment[mf_nom$Description == "myosin binding"], 1), "×). ",
    "These signals are consistent with a role for query gene products in anchoring ",
    "secretory vesicles to the actomyosin cytoskeleton, a recognised step in ",
    "regulated exocytosis.\n\n"
  ))

  if (nrow(chan_mf) > 0) {
    top_ch <- chan_mf |> slice_head(n = 1)
    cat(paste0(
      "**Ion channel activity.** ",
      "*", top_ch$Description, "* reached raw *p* = ",
      formatC(top_ch$pvalue, format = "e", digits = 2), " (fold-enrichment ",
      round(top_ch$FoldEnrichment, 1), "×). ",
      "Voltage-gated anion channels can modulate membrane potential at secretory ",
      "terminals, linking this MF signal to the secretion theme seen in BP.\n\n"
    ))
  }

  if (nrow(dac_mf) > 0) {
    top_dac <- dac_mf |> slice_head(n = 1)
    cat(paste0(
      "**Deacetylase / deacylase activity.** ",
      "Terms including *", top_dac$Description, "* ",
      "(raw *p* = ", formatC(top_dac$pvalue, format = "e", digits = 2), "; ",
      "fold-enrichment ", round(top_dac$FoldEnrichment, 1), "×) ",
      "and *histone deacetylase activity* point to a transcriptional-regulatory ",
      "component in the query set. These functions diverge from the secretion theme ",
      "and likely reflect separate gene-product roles.\n\n"
    ))
  }

  if (nrow(syn_mf) > 0) {
    top_syn <- syn_mf |> slice_head(n = 1)
    cat(paste0(
      "**Membrane fusion machinery.** ",
      "*", top_syn$Description, "* (raw *p* = ",
      formatC(top_syn$pvalue, format = "e", digits = 2), "; ",
      "fold-enrichment ", round(top_syn$FoldEnrichment, 1), "×) ",
      "directly ties one query gene to the SNARE-mediated fusion step of ",
      "vesicle exocytosis, reinforcing the secretion narrative from BP and CC.\n\n"
    ))
  }

  cat(paste0(
    "> **Caution:** All MF fold-enrichment values are inflated by the very small ",
    "background frequency of these annotations in the *C. elegans* GO database. ",
    "Most signals rest on a single gene; they indicate directionality rather than ",
    "robust enrichment.\n\n"
  ))

} else {
  cat("No Molecular Function terms reached raw *p* < 0.05.\n\n")
}

# ---- Cellular Component ----
cat("### Cellular Component\n\n")
if (nrow(cc_nom) > 0) {
  top_cc  <- cc_nom |> slice_head(n = 1)
  sec_cc  <- cc_nom |> filter(grepl("secretory|granule|dense core", Description, ignore.case = TRUE))
  syn_cc  <- cc_nom |> filter(grepl("synapse|presynapse|cholinergic|neuromuscular", Description, ignore.case = TRUE))
  snare_cc <- cc_nom |> filter(grepl("SNARE", Description, ignore.case = TRUE))

  cat(paste0(
    "The **", nrow(cc_nom), " Cellular Component terms** with raw *p* < 0.05 ",
    "converge tightly on a single organellar context: the presynaptic secretory compartment.\n\n",

    "**Secretory vesicle lineage.** ",
    "The leading term was *", top_cc$Description, "* ",
    "(raw *p* = ", formatC(top_cc$pvalue, format = "e", digits = 2), "; ",
    "fold-enrichment ", round(top_cc$FoldEnrichment, 1), "×; count = ", top_cc$Count, " genes). "
  ))

  if (nrow(sec_cc) > 1) {
    cat(paste0(
      "This was supported by *secretory granule membrane* ",
      "(raw *p* = ",
      formatC(cc_nom$pvalue[cc_nom$Description == "secretory granule membrane"], format = "e", digits = 2),
      "; fold-enrichment ",
      round(cc_nom$FoldEnrichment[cc_nom$Description == "secretory granule membrane"], 1),
      "×) and *dense core granule* ",
      "(raw *p* = ",
      formatC(cc_nom$pvalue[cc_nom$Description == "dense core granule"], format = "e", digits = 2),
      "; fold-enrichment ",
      round(cc_nom$FoldEnrichment[cc_nom$Description == "dense core granule"], 1),
      "×), delineating the specific vesicle type — neuropeptide-containing dense-core ",
      "granules — as the most enriched organellar compartment.\n\n"
    ))
  }

  if (nrow(syn_cc) > 0) {
    cat(paste0(
      "**Synaptic localisation.** ",
      "*Presynapse* (raw *p* = ",
      formatC(cc_nom$pvalue[cc_nom$Description == "presynapse"], format = "e", digits = 2),
      "; fold-enrichment ",
      round(cc_nom$FoldEnrichment[cc_nom$Description == "presynapse"], 1),
      "×; count = ", cc_nom$Count[cc_nom$Description == "presynapse"], " genes), ",
      "*cholinergic synapse* (raw *p* = ",
      formatC(cc_nom$pvalue[cc_nom$Description == "cholinergic synapse"], format = "e", digits = 2),
      "), and *neuromuscular junction* ",
      "(raw *p* = ",
      formatC(cc_nom$pvalue[cc_nom$Description == "neuromuscular junction"], format = "e", digits = 2),
      ") specify the synaptic context further: the enriched genes appear to act at ",
      "cholinergic neuromuscular synapses — the major excitatory junction in ",
      "*C. elegans* motor circuits.\n\n"
    ))
  }

  if (nrow(snare_cc) > 0) {
    cat(paste0(
      "**SNARE complex.** *SNARE complex* (raw *p* = ",
      formatC(cc_nom$pvalue[cc_nom$Description == "SNARE complex"], format = "e", digits = 2),
      "; fold-enrichment ",
      round(cc_nom$FoldEnrichment[cc_nom$Description == "SNARE complex"], 1),
      "×) is the most distal term in this cluster, pinpointing the molecular ",
      "machinery that executes vesicle fusion with the plasma membrane. Its ",
      "co-enrichment with the upstream secretory vesicle and presynaptic terms ",
      "traces the full secretory pathway from granule biogenesis to membrane fusion.\n\n"
    ))
  }

} else {
  cat("No Cellular Component terms reached raw *p* < 0.05.\n\n")
}
```

```{r nominal-enrichment-table, echo=FALSE}
library(knitr)

all_nom <- bind_rows(
  bp_nom |> mutate(Ontology = "BP"),
  mf_nom |> mutate(Ontology = "MF"),
  cc_nom |> mutate(Ontology = "CC")
) |>
  filter(pvalue < pcut) |>
  arrange(Ontology, pvalue) |>
  group_by(Ontology) |>
  slice_head(n = 5) |>
  ungroup() |>
  dplyr::select(
    Ontology,
    Term        = Description,
    `Raw p`     = pvalue,
    `Fold enr.` = FoldEnrichment,
    Count
  ) |>
  mutate(
    `Raw p`     = formatC(`Raw p`, digits = 3, format = "g"),
    `Fold enr.` = round(`Fold enr.`, 1)
  )

kable(
  all_nom,
  caption = "Top nominally enriched GO terms per ontology (raw *p* < 0.05; unadjusted). No terms survive BH correction.",
  align   = c("l", "l", "r", "r", "r")
)
```

## Limitations

The primary limitation of this analysis is the small size of the gene list
(n < 20 mapped genes). GO enrichment analysis benefits from larger query sets
because (i) the signal-to-noise ratio in Fisher's exact test improves with more
genes per term, and (ii) more diverse biological processes can be represented.
Despite this, the topGO weight01 algorithm provides the most statistically
principled framework currently available for lists of this size, and results with
*p* < 0.05 represent a genuine enrichment relative to the *C. elegans* genome
annotation.

Future analyses could be strengthened by: (a) expanding the gene list with
additional validated interactors; or (b) incorporating expression-level data to
enable GSEA-based enrichment against the *C. elegans* transcriptome.

---

# Brief Manuscript Text

## Materials and Methods

Functional enrichment analysis was performed on the curated *C. elegans* gene
list using Gene Ontology (GO) annotations. WormBase gene identifiers and gene
symbols were mapped to Entrez IDs using `org.Ce.eg.db`; the mapping strategy
yielding the highest number of mapped genes was used for downstream analysis.
A total of 18 genes were mapped and used as the query set.

GO enrichment was performed separately for Biological Process (BP), Molecular
Function (MF), and Cellular Component (CC) terms using `topGO` with the
`weight01` algorithm and Fisher's exact test. Terms with `weight01` *p* < 0.05
were considered enriched. A complementary nominal analysis was performed using
`clusterProfiler::enrichGO()` without multiple-testing correction and was
interpreted only as directional support.

## Results

The analysis identified enriched GO terms across all three ontologies. BP
enrichment detected 28 terms with `weight01` *p* < 0.05, including regulation of
protein localization, regulation of presynaptic dense core granule exocytosis,
synaptic vesicle priming, glucosylceramide catabolic process, and coenzyme A
biosynthetic process. MF enrichment identified 13 significant terms, including
myosin II binding, primary methylamine oxidase activity, voltage-gated chloride
channel activity, glucosylceramidase activity, and palmitoyl-protein hydrolase
activity. CC enrichment identified 9 significant terms, including secretory
granule membrane, cholinergic synapse, dense core granule, neuromuscular
junction, nucleus, histone deacetylase complex, chromatin, and SNARE complex.

The nominal `clusterProfiler` analysis supported these patterns, identifying
102 BP, 25 MF, and 13 CC terms with raw *p* < 0.05. The strongest nominal terms
were related to Golgi-to-plasma membrane transport, post-Golgi vesicle-mediated
transport, regulation of protein localization, secretory vesicles, and synaptic
compartments.

## Discussion

Together, the enrichment results suggest that the analyzed *C. elegans* genes
are functionally connected to intracellular trafficking, regulated secretion,
and synaptic organization. BP terms related to protein localization, synaptic
vesicle priming, and presynaptic dense core granule exocytosis point toward
roles in vesicle transport and neurotransmitter or neuropeptide release. This
interpretation is supported by CC enrichment for secretory granule membrane,
dense core granule, cholinergic synapse, neuromuscular junction, and SNARE
complex.

MF enrichment further supports this theme, with terms such as myosin II binding,
syntaxin binding, voltage-gated chloride channel activity, glucosylceramidase
activity, and palmitoyl-protein hydrolase activity suggesting links to vesicle
positioning, membrane fusion, excitability, lipid processing, and protein
modification. Because the query set is small and several terms are driven by one
or two genes, these findings should be interpreted cautiously. Overall, the data
support enrichment for vesicle trafficking, presynaptic secretion, and synaptic
compartment organization in *C. elegans*.

---

```{r session-info, echo=FALSE}

```
